---
title: "ETO5513_T2DM_Risk-Factors"
author: "Eileen Koh"
date: "2026-04-01"
output:
  bookdown::html_document2:
    keep_md: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Libraries
library(tidyverse)
library(knitr)
library(ggplot2)

{r, include=TRUE}
# Load dataset 
data <- read_csv("Data/diabetes_dataset.csv")

# Standardize
data <- data %>%
  mutate(
    diabetes = ifelse(diagnosed_diabetes == 1, "Yes", "No"),
    education_level = case_when(
      education_level %in% c("Post Graduate", "Postgraduate") ~ "Postgraduate",
      education_level %in% c("No Formal", "No formal") ~ "No formal",
      TRUE ~ education_level
    ),
    income_level = case_when(
      income_level %in% c("Lower middle", "Lower-Middle") ~ "Lower-Middle",
      income_level %in% c("Upper middle", "Upper-Middle") ~ "Upper-Middle",
      TRUE ~ income_level
    )
  )


fmt <- function(x, total) {
  paste0(x, " (", round(x/total*100,1), "%)")
}

# Split groups
yes <- data %>% filter(diabetes == "Yes")
no  <- data %>% filter(diabetes == "No")

# Totals
n_yes <- nrow(yes)
n_no  <- nrow(no)

# Employment levels (ordered manually)
employment_levels <- c("Employed", "Unemployed", "Student", "Retired")

# Build table (FIXED)
table1 <- data.frame(
  Variable = c(
    "Age (Mean ± SD)", "",
    
    "Gender",
    "  Male",
    "  Female",
    "",
    
    "Ethnicity",
    "  Asian",
    "  White",
    "  Hispanic",
    "  Black",
    "  Other",
    "",
    
    "Education Level",
    "  Highschool",
    "  Graduate",
    "  Postgraduate",
    "  No formal",
    "",
    
    "Income Level",
    "  Low",
    "  Lower-Middle",
    "  Middle",
    "  Upper-Middle",
    "  High",
    "",
    
    "Employment Status",
    "  Employed",
    "  Unemployed",
    "  Student",
    "  Retired"
  ),
  
  `No Diabetes` = c(
    paste0(round(mean(no$age, na.rm=TRUE),1), " ± ", round(sd(no$age, na.rm=TRUE),1)),
    "",
    
    "",
    fmt(sum(no$gender=="Male", na.rm=TRUE), n_no),
    fmt(sum(no$gender=="Female", na.rm=TRUE), n_no),
    "",
    
    "",
    fmt(sum(no$ethnicity=="Asian", na.rm=TRUE), n_no),
    fmt(sum(no$ethnicity=="White", na.rm=TRUE), n_no),
    fmt(sum(no$ethnicity=="Hispanic", na.rm=TRUE), n_no),
    fmt(sum(no$ethnicity=="Black", na.rm=TRUE), n_no),
    fmt(sum(no$ethnicity=="Other", na.rm=TRUE), n_no),
    "",
    
    "",
    fmt(sum(no$education_level=="Highschool", na.rm=TRUE), n_no),
    fmt(sum(no$education_level=="Graduate", na.rm=TRUE), n_no),
    fmt(sum(no$education_level=="Postgraduate", na.rm=TRUE), n_no),
    fmt(sum(no$education_level=="No formal", na.rm=TRUE), n_no),
    "",
    
    "",
    fmt(sum(no$income_level=="Low", na.rm=TRUE), n_no),
    fmt(sum(no$income_level=="Lower-Middle", na.rm=TRUE), n_no),
    fmt(sum(no$income_level=="Middle", na.rm=TRUE), n_no),
    fmt(sum(no$income_level=="Upper-Middle", na.rm=TRUE), n_no),
    fmt(sum(no$income_level=="High", na.rm=TRUE), n_no),
    "",
    
    "",
    fmt(sum(no$employment_status=="Employed", na.rm=TRUE), n_no),
    fmt(sum(no$employment_status=="Unemployed", na.rm=TRUE), n_no),
    fmt(sum(no$employment_status=="Student", na.rm=TRUE), n_no),
    fmt(sum(no$employment_status=="Retired", na.rm=TRUE), n_no)
  ),
  
  `Diabetes` = c(
    paste0(round(mean(yes$age, na.rm=TRUE),1), " ± ", round(sd(yes$age, na.rm=TRUE),1)),
    "",
    
    "",
    fmt(sum(yes$gender=="Male", na.rm=TRUE), n_yes),
    fmt(sum(yes$gender=="Female", na.rm=TRUE), n_yes),
    "",
    
    "",
    fmt(sum(yes$ethnicity=="Asian", na.rm=TRUE), n_yes),
    fmt(sum(yes$ethnicity=="White", na.rm=TRUE), n_yes),
    fmt(sum(yes$ethnicity=="Hispanic", na.rm=TRUE), n_yes),
    fmt(sum(yes$ethnicity=="Black", na.rm=TRUE), n_yes),
    fmt(sum(yes$ethnicity=="Other", na.rm=TRUE), n_yes),
    "",
    
    "",
    fmt(sum(yes$education_level=="Highschool", na.rm=TRUE), n_yes),
    fmt(sum(yes$education_level=="Graduate", na.rm=TRUE), n_yes),
    fmt(sum(yes$education_level=="Postgraduate", na.rm=TRUE), n_yes),
    fmt(sum(yes$education_level=="No formal", na.rm=TRUE), n_yes),
    "",
    
    "",
    fmt(sum(yes$income_level=="Low", na.rm=TRUE), n_yes),
    fmt(sum(yes$income_level=="Lower-Middle", na.rm=TRUE), n_yes),
    fmt(sum(yes$income_level=="Middle", na.rm=TRUE), n_yes),
    fmt(sum(yes$income_level=="Upper-Middle", na.rm=TRUE), n_yes),
    fmt(sum(yes$income_level=="High", na.rm=TRUE), n_yes),
    "",
    
    "",
    fmt(sum(yes$employment_status=="Employed", na.rm=TRUE), n_yes),
    fmt(sum(yes$employment_status=="Unemployed", na.rm=TRUE), n_yes),
    fmt(sum(yes$employment_status=="Student", na.rm=TRUE), n_yes),
    fmt(sum(yes$employment_status=="Retired", na.rm=TRUE), n_yes)
  ),
  
  stringsAsFactors = FALSE
)

kable(table1, caption = "Table 1: Demographics by Diabetes Diagnosis", align = "lcc")
```
Individuals with diabetes are on average, older than those without diabetes (51.9 vs 47.5 years), indicating age may be associated with increased likelihood of diabetes diagnosis. In contrast, all other demographic characteristics are highly similar between the two groups. Gender distribution is nearly identical, with males and females each comprising roughly half of both populations. Ethnicity is also consistent, with White individuals representing the largest proportion (~45%), followed by Hispanic, Black, Asian, and Other groups. Education, income, and employment status show minimal variation, suggesting no substantial socio-demographic differences apart from age in this dataset.

```{r, include=TRUE}
# Means
mean_yes <- mean(yes$age, na.rm = TRUE)
mean_no  <- mean(no$age, na.rm = TRUE)
# Distribution
ggplot(data, aes(x = age, fill = diabetes)) +
  geom_density(alpha = 0.5, colour = NA) +
  
  # Mean lines
  geom_vline(xintercept = mean_no, linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = mean_yes, linetype = "dashed", linewidth = 0.8) +
  
  # Labels
  annotate("text", x = mean_no, y = 0.025,
           label = paste0("No: ", round(mean_no,1)),
           angle = 90, vjust = -0.5, size = 3) +
  
  annotate("text", x = mean_yes, y = 0.025,
           label = paste0("Yes: ", round(mean_yes,1)),
           angle = 90, vjust = -0.5, size = 3) +
  
  
  labs(
    title = "Chart 1: Age Distribution by Diabetes Status",
    subtitle = "Density plot with mean age indicated",
    x = "Age (years)",
    y = "Density",
    fill = "Diabetes Status"
  ) +
  
  # Tufte-style minimalism
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 10),
    panel.grid.minor = element_blank()
  )
```
Individuals with diabetes were older on average than those without the condition. The density curves showed a clear skew towards the right for the diabetic group, with mean age higher (51.1 years) compared with non‑diabetic participants (47.4 years). The diabetic distribution was more concentrated around middle age, while the non‑diabetic group displayed a broader spread. These patterns indicate a consistent association between increasing age and diabetes presence.

```{r, include=TRUE}

# Grouping
data <- data %>%
  mutate(
    physical_activity_group = ifelse(physical_activity_minutes_per_week >= median(physical_activity_minutes_per_week, na.rm = TRUE),
                                     "High Activity", "Low Activity"),

    diet_group = ifelse(diet_score >= median(diet_score, na.rm = TRUE),
                        "Good Diet", "Poor Diet"),

    sleep_group = ifelse(sleep_hours_per_day >= 7,
                         "Adequate Sleep", "Low Sleep"),

    screen_time_group = ifelse(screen_time_hours_per_day >= median(screen_time_hours_per_day, na.rm = TRUE),
                               "High Screen Time", "Low Screen Time")
  )

# Mean ± SD
mean_sd <- function(x) {
  paste0(round(mean(x, na.rm = TRUE), 1),
         " ± ",
         round(sd(x, na.rm = TRUE), 1))
}

# Table 2
table2 <- data.frame(
  Variable = c(
    "Physical Activity",
    "Low Activity",
    "High Activity",
    "",
    
    "Diet Score",
    "Poor Diet",
    "Good Diet",
    "",
    
    "Sleep",
    "Low Sleep",
    "Adequate Sleep",
    "",
    
    "Screen Time",
    "Low Screen Time",
    "High Screen Time"
  ),

  `Diabetes Risk Score` = c(
    "",
    mean_sd(data$diabetes_risk_score[data$physical_activity_group == "Low Activity"]),
    mean_sd(data$diabetes_risk_score[data$physical_activity_group == "High Activity"]),
    "",
    
    "",
    mean_sd(data$diabetes_risk_score[data$diet_group == "Poor Diet"]),
    mean_sd(data$diabetes_risk_score[data$diet_group == "Good Diet"]),
    "",
    
    "",
    mean_sd(data$diabetes_risk_score[data$sleep_group == "Low Sleep"]),
    mean_sd(data$diabetes_risk_score[data$sleep_group == "Adequate Sleep"]),
    "",
    
    "",
    mean_sd(data$diabetes_risk_score[data$screen_time_group == "Low Screen Time"]),
    mean_sd(data$diabetes_risk_score[data$screen_time_group == "High Screen Time"])
  )
)

kable(table2,
      caption = "Table 2: Diabetes Risk Score by Lifestyle Factors",
      align = "lc")
```
Table 2 shows clear variation in diabetes risk score across lifestyle groups. Individuals with low physical activity have a higher mean risk score (~33–35) compared to those with high activity (~20–25). A similar pattern is seen for diet, where a poor diet group records a higher risk score (~30–34) than the good diet group (~22–26). For sleep, those with low sleep show greater risk (~32–36) compared to adequate sleep (~24–28). Finally, high screen time is associated with higher risk (~31–35) versus low screen time (~22–26). Overall, healthier behaviours consistently correspond to lower diabetes risk.

```{r, include=TRUE}
ggplot(data, aes(screen_time_hours_per_day, diabetes_risk_score)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  theme_minimal() +
  labs(
    title = "Chart 2: Scatter Plot Screen Time vs Diabetes Risk",
    x = "Screen Time (hours per day)",
    y = "Diabetes Risk Score"
  )
```

Chart 2 shows the relationship between screen time and diabetes risk score. The scatterplot reveals a highly dispersed distribution of points, indicating substantial variability in risk scores across all levels of screen time. The fitted regression line shows a very weak positive trend, suggesting that higher screen time is only marginally associated with increased diabetes risk. Overall, screen time alone appears to be a poor predictor of diabetes risk.
