---
title: "ETO5513_T2DM_Risk-Factors"
author: 
  - "Kale-Ab Bobe"
  - "Eileen Koh"
date: "2026-04-01"
output:
  bookdown::html_document2:
    keep_md: true
---


# Introduction

According to the 11th edition of the IDF Diabetes Atlas, more than one in eight adults worldwide will be living with diabetes by 2050, and the global burden is projected to rise sharply in the coming years (International Diabetes Federation, 2025). The complications of diabetes include higher risks of cardiovascular disease, stroke, heart attack, and heart failure (Dam, Verma, & Jasrotia, 2025). Obesity, dietary risks, occupational risks, tobacco use, low physical activity, and alcohol consumption are major risk factors for type 2 diabetes (Ong, Stafford, McLaughlin, et al., 2023).
Emerging evidence also suggests that sedentary behaviours, such as prolonged screen time, and inadequate or excessive sleep duration may contribute to increased diabetes risk through their effects on metabolic regulation and insulin sensitivity (Park, Moon, Kim, Kong, & Oh, 2020). Clinical biomarkers such as glycated haemoglobin (HbA1c), blood pressure, and cholesterol are widely recognized as key indicators of metabolic health, strongly linked to diabetes diagnosis, progression, and cardiovascular risk. Evaluating the interplay between these parameters and diabetes status yields valuable insights into disease management, risk stratification, and long-term health outcomes (Savvopoulos, Hatzikirou, & Jelinek, 2024).
This study aims to: (1) describe the demographic profile of individuals with and without diabetes; (2) assess the association between lifestyle factors and diabetes; (3) evaluate the relationship between clinical variables and diabetes; and (4) identify the overall key risk factors for diabetes using logistic regression (5) identify critical thresholds for risk classification.


# Analysis sections



```{r load_libraries, echo=FALSE, message=FALSE, warning=FALSE}


library(knitr)
library(markdown)
library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(bookdown)
library(car)
library(broom)
library(tableone)
```
## Demographics by Diabetes Diagnosis

```{r table1, echo=FALSE, show_col_types = FALSE}
# Load dataset 
data <- read_csv("Data/diabetes_dataset.csv")

# Standardize
data <- data %>%
  mutate(
    diabetes = ifelse(diagnosed_diabetes == 1, "Yes", "No"),
    education_level = case_when(
      education_level %in% c("Post Graduate", "Postgraduate") ~ "Postgraduate",
      education_level %in% c("No Formal", "No formal") ~ "No formal",
      TRUE ~ education_level
    ),
    income_level = case_when(
      income_level %in% c("Lower middle", "Lower-Middle") ~ "Lower-Middle",
      income_level %in% c("Upper middle", "Upper-Middle") ~ "Upper-Middle",
      TRUE ~ income_level
    )
  )


fmt <- function(x, total) {
  paste0(x, " (", round(x/total*100,1), "%)")
}

# Split groups
yes <- data %>% filter(diabetes == "Yes")
no  <- data %>% filter(diabetes == "No")

# Totals
n_yes <- nrow(yes)
n_no  <- nrow(no)

# Employment levels
employment_levels <- c("Employed", "Unemployed", "Student", "Retired")

# Table
table1 <- data.frame(
  Variable = c(
    "Age (Mean ± SD)", "",
    
    "Gender",
    "  Male",
    "  Female",
    "",
    
    "Ethnicity",
    "  Asian",
    "  White",
    "  Hispanic",
    "  Black",
    "  Other",
    "",
    
    "Education Level",
    "  Highschool",
    "  Graduate",
    "  Postgraduate",
    "  No formal",
    "",
    
    "Income Level",
    "  Low",
    "  Lower-Middle",
    "  Middle",
    "  Upper-Middle",
    "  High",
    "",
    
    "Employment Status",
    "  Employed",
    "  Unemployed",
    "  Student",
    "  Retired"
  ),
  
  `No Diabetes` = c(
    paste0(round(mean(no$age, na.rm=TRUE),1), " ± ", round(sd(no$age, na.rm=TRUE),1)),
    "",
    
    "",
    fmt(sum(no$gender=="Male", na.rm=TRUE), n_no),
    fmt(sum(no$gender=="Female", na.rm=TRUE), n_no),
    "",
    
    "",
    fmt(sum(no$ethnicity=="Asian", na.rm=TRUE), n_no),
    fmt(sum(no$ethnicity=="White", na.rm=TRUE), n_no),
    fmt(sum(no$ethnicity=="Hispanic", na.rm=TRUE), n_no),
    fmt(sum(no$ethnicity=="Black", na.rm=TRUE), n_no),
    fmt(sum(no$ethnicity=="Other", na.rm=TRUE), n_no),
    "",
    
    "",
    fmt(sum(no$education_level=="Highschool", na.rm=TRUE), n_no),
    fmt(sum(no$education_level=="Graduate", na.rm=TRUE), n_no),
    fmt(sum(no$education_level=="Postgraduate", na.rm=TRUE), n_no),
    fmt(sum(no$education_level=="No formal", na.rm=TRUE), n_no),
    "",
    
    "",
    fmt(sum(no$income_level=="Low", na.rm=TRUE), n_no),
    fmt(sum(no$income_level=="Lower-Middle", na.rm=TRUE), n_no),
    fmt(sum(no$income_level=="Middle", na.rm=TRUE), n_no),
    fmt(sum(no$income_level=="Upper-Middle", na.rm=TRUE), n_no),
    fmt(sum(no$income_level=="High", na.rm=TRUE), n_no),
    "",
    
    "",
    fmt(sum(no$employment_status=="Employed", na.rm=TRUE), n_no),
    fmt(sum(no$employment_status=="Unemployed", na.rm=TRUE), n_no),
    fmt(sum(no$employment_status=="Student", na.rm=TRUE), n_no),
    fmt(sum(no$employment_status=="Retired", na.rm=TRUE), n_no)
  ),
  
  `Diabetes` = c(
    paste0(round(mean(yes$age, na.rm=TRUE),1), " ± ", round(sd(yes$age, na.rm=TRUE),1)),
    "",
    
    "",
    fmt(sum(yes$gender=="Male", na.rm=TRUE), n_yes),
    fmt(sum(yes$gender=="Female", na.rm=TRUE), n_yes),
    "",
    
    "",
    fmt(sum(yes$ethnicity=="Asian", na.rm=TRUE), n_yes),
    fmt(sum(yes$ethnicity=="White", na.rm=TRUE), n_yes),
    fmt(sum(yes$ethnicity=="Hispanic", na.rm=TRUE), n_yes),
    fmt(sum(yes$ethnicity=="Black", na.rm=TRUE), n_yes),
    fmt(sum(yes$ethnicity=="Other", na.rm=TRUE), n_yes),
    "",
    
    "",
    fmt(sum(yes$education_level=="Highschool", na.rm=TRUE), n_yes),
    fmt(sum(yes$education_level=="Graduate", na.rm=TRUE), n_yes),
    fmt(sum(yes$education_level=="Postgraduate", na.rm=TRUE), n_yes),
    fmt(sum(yes$education_level=="No formal", na.rm=TRUE), n_yes),
    "",
    
    "",
    fmt(sum(yes$income_level=="Low", na.rm=TRUE), n_yes),
    fmt(sum(yes$income_level=="Lower-Middle", na.rm=TRUE), n_yes),
    fmt(sum(yes$income_level=="Middle", na.rm=TRUE), n_yes),
    fmt(sum(yes$income_level=="Upper-Middle", na.rm=TRUE), n_yes),
    fmt(sum(yes$income_level=="High", na.rm=TRUE), n_yes),
    "",
    
    "",
    fmt(sum(yes$employment_status=="Employed", na.rm=TRUE), n_yes),
    fmt(sum(yes$employment_status=="Unemployed", na.rm=TRUE), n_yes),
    fmt(sum(yes$employment_status=="Student", na.rm=TRUE), n_yes),
    fmt(sum(yes$employment_status=="Retired", na.rm=TRUE), n_yes)
  ),
  
  stringsAsFactors = FALSE
)

kable(table1, caption = "Demographics by Diabetes Diagnosis", align = "lcc")
```


Interpretation: Table \@ref(tab:table1) indicates that individuals with diabetes are, on average, older than those without (51.9 vs 47.5 years), suggesting age is associated with a higher likelihood of diagnosis. This aligns with existing evidence linking increasing age to greater diabetes risk due to metabolic and physiological changes over time (International Diabetes Federation, 2025; Hu et al., 2020). In contrast, other demographic characteristics including gender, ethnicity, education, income, and employment show minimal variation between groups.


## Age Distribution by Diabetes Status


```{r chart1, echo=FALSE, fig.cap="Age Distribution by Diabetes Status"}
# Means
mean_yes <- mean(yes$age, na.rm = TRUE)
mean_no  <- mean(no$age, na.rm = TRUE)
# Distribution
ggplot(data, aes(x = age, fill = diabetes)) +
  geom_density(alpha = 0.5, colour = NA) +
  
  # Mean lines
  geom_vline(xintercept = mean_no, linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = mean_yes, linetype = "dashed", linewidth = 0.8) +
  
  # Labels
  annotate("text", x = mean_no, y = 0.025,
           label = paste0("No: ", round(mean_no,1)),
           angle = 90, vjust = -0.5, size = 3) +
  
  annotate("text", x = mean_yes, y = 0.025,
           label = paste0("Yes: ", round(mean_yes,1)),
           angle = 90, vjust = -0.5, size = 3) +
  
  
  labs(
    title = "Age Distribution by Diabetes Status",
    subtitle = "Density plot with mean age indicated",
    x = "Age (years)",
    y = "Density",
    fill = "Diabetes Status"
  ) +
  
  # Tufte-style minimalism
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 10),
    panel.grid.minor = element_blank()
  )



```


Interpretation: Chart \@ref(fig:chart1) reinforces the pattern observed in Table @ref(tab:table1), with the diabetic group displaying a right-skewed age distribution and a higher mean (51.1 years) age compared to non-diabetic individuals (47.4 years). This further supports the role of age as a consistent risk factor (Hu et al., 2020). The narrower concentration around middle age among diabetic individuals also suggests a critical period where risk becomes more pronounced, potentially reflecting cumulative exposure to metabolic and lifestyle risk factors over time.


## Diabetes Risk Score by Lifestyle Factors

```{r table2, echo=FALSE}

# Grouping
data <- data %>%
  mutate(
    physical_activity_group = ifelse(physical_activity_minutes_per_week >= median(physical_activity_minutes_per_week, na.rm = TRUE),
                                     "High Activity", "Low Activity"),

    diet_group = ifelse(diet_score >= median(diet_score, na.rm = TRUE),
                        "Good Diet", "Poor Diet"),

    sleep_group = ifelse(sleep_hours_per_day >= 7,
                         "Adequate Sleep", "Low Sleep"),

    screen_time_group = ifelse(screen_time_hours_per_day >= median(screen_time_hours_per_day, na.rm = TRUE),
                               "High Screen Time", "Low Screen Time")
  )



# Mean ± SD
mean_sd <- function(x) {
  paste0(round(mean(x, na.rm = TRUE), 1),
         " ± ",
         round(sd(x, na.rm = TRUE), 1))
}

# Table 2
table2 <- data.frame(
  Variable = c(
    "Physical Activity",
    "Low Activity",
    "High Activity",
    "",
    
    "Diet Score",
    "Poor Diet",
    "Good Diet",
    "",
    
    "Sleep",
    "Low Sleep",
    "Adequate Sleep",
    "",
    
    "Screen Time",
    "Low Screen Time",
    "High Screen Time"
  ),

  `Diabetes Risk Score` = c(
    "",
    mean_sd(data$diabetes_risk_score[data$physical_activity_group == "Low Activity"]),
    mean_sd(data$diabetes_risk_score[data$physical_activity_group == "High Activity"]),
    "",
    
    "",
    mean_sd(data$diabetes_risk_score[data$diet_group == "Poor Diet"]),
    mean_sd(data$diabetes_risk_score[data$diet_group == "Good Diet"]),
    "",
    
    "",
    mean_sd(data$diabetes_risk_score[data$sleep_group == "Low Sleep"]),
    mean_sd(data$diabetes_risk_score[data$sleep_group == "Adequate Sleep"]),
    "",
    
    "",
    mean_sd(data$diabetes_risk_score[data$screen_time_group == "Low Screen Time"]),
    mean_sd(data$diabetes_risk_score[data$screen_time_group == "High Screen Time"])
  )
)

kable(table2,
      caption = "Diabetes Risk Score by Lifestyle Factors",
      align = "lc")
```

Interpretation: Table \@ref(tab:table2) shows clear variation in diabetes risk score across lifestyle groups. Individuals with low physical activity have a higher mean risk score (~33–35) compared to those with high activity (~20–25). A similar pattern is seen for diet, where a poor diet group records a higher risk score (~30–34) than the good diet group (~22–26). For sleep, those with low sleep show greater risk (~32–36) compared to adequate sleep (~24–28). Finally, high screen time is associated with higher risk (~31–35) versus low screen time (~22–26). Overall, healthier behaviours consistently correspond to lower diabetes risk, consistent with prior research on behavioural risk factors (Ong et al., 2023; Park et al., 2020). 


## Scatter Plot Screen Time vs Diabetes Risk

```{r chart2, echo=FALSE, fig.cap="Scatter Plot Screen Time vs Diabetes Risk"}
ggplot(data, aes(screen_time_hours_per_day, diabetes_risk_score)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  theme_minimal() +
  labs(
    title = "Scatter Plot Screen Time vs Diabetes Risk",
    x = "Screen Time (hours per day)",
    y = "Diabetes Risk Score"
  )
```

Interpretation:Chart \@ref(fig:chart2) shows that screen time alone has only a weak positive association with diabetes risk, suggesting it is not a strong independent predictor despite its broader behavioural relevance (Park et al., 2020). This indicates that screen time alone doesn’t directly influence risk of diabetes.



## Association of Clinical Variables with Diabetes Diagnosis 
To understand the association of clinical variables with diabetes, independent t-test and chi-square test are performed to determine the potential significant risk factors of diabetes.

```{r table3, echo=FALSE}

data <- data %>%
  mutate(
    diagnosed_diabetes = factor(diagnosed_diabetes, levels = c(0, 1), labels = c("No Diabetes", "Diabetes")),
    family_history_diabetes = factor(family_history_diabetes, levels = c(0, 1), labels = c("No", "Yes")),
    hypertension_history = factor(hypertension_history, levels = c(0, 1), labels = c("No", "Yes")),
    cardiovascular_history = factor(cardiovascular_history, levels = c(0, 1), labels = c("No", "Yes"))
  )
  
vars <- c( "bmi", "waist_to_hip_ratio", "systolic_bp",
           "diastolic_bp", "heart_rate", "cholesterol_total", "hdl_cholesterol",
           "ldl_cholesterol", "triglycerides", "glucose_fasting", "glucose_postprandial", "hba1c", "diabetes_risk_score",
           "diabetes_stage")
table_three <- CreateTableOne(vars = vars,
                              strata = "diagnosed_diabetes",
                              data = data,
                              factorVars = c( "diabetes_stage"))



knitr::kable(
  print(table_three, showAllLevels = TRUE),
  caption = "Baseline Clinical Parameters of Patients with and without Diabetes"
)



```

Interpretation:

Table \@ref(tab:table3) reveals the differences in clinical parameters between individuals with and without diagnosed diabetes. Among 100,000 participants, 40,002 were non-diabetic and 59,998 diabetic. Those with diabetes showed higher BMI, waist-to-hip ratio, blood pressure, heart rate, and adverse lipid profiles, including elevated total cholesterol, LDL, and triglycerides, alongside lower HDL. Glucose measures were also higher, with fasting, postprandial, and HbA1c values well above non-diabetic levels. Nearly all diagnosed cases were Type 2 diabetes (99.6%), while most non-diabetics were classified as pre-diabetes. All differences were statistically significant (p < 0.001).



## Boxplot of Systolic BP by Diabetes Diagnosis

```{r chart3, echo=FALSE, fig.cap="Boxplot of Systolic BP by Diabetes Diagnosis"}

ggplot(data, aes(x = diagnosed_diabetes, y = systolic_bp, fill = diagnosed_diabetes)) +
  geom_boxplot(alpha = 0.5) +
  scale_fill_manual(values = c("purple", "green")) +
  labs(title = "Systolic BP by Diabetes Diagnosis",
       y = "Systolic BP", x = "Diabetes Diagnosis") +
  theme_minimal() +
  theme(legend.position = "none")
```

Interpretation:


Chart \@ref(fig:chart3) compares systolic blood pressure between individuals with and without diabetes. The median systolic BP is slightly higher among those diagnosed with diabetes, with a broader distribution and more outliers.


To determine the risk factors of diabetes, potential factors from the above analysis will be included in the logistic regression model to obtain the adjusted coefficients. 



## Risk Factors of Diabetes using Logistic Regression 


```{r LR, echo=FALSE}

# Run logistic regression

model <- glm(diagnosed_diabetes ~ age  + family_history_diabetes + hypertension_history +cardiovascular_history + screen_time_hours_per_day + bmi + physical_activity_minutes_per_week  + diet_score + hba1c + glucose_fasting + glucose_postprandial + waist_to_hip_ratio + systolic_bp + diastolic_bp + heart_rate + cholesterol_total + hdl_cholesterol +  ldl_cholesterol + triglycerides, 
             data = data,
             family = binomial)
vif(model)
summary(model)

# Obtain the odds ratio

table_or <- tidy(model) %>%
  mutate(
    OR = exp(estimate),
    CI_low = exp(estimate - 1.96 * std.error),
    CI_high = exp(estimate + 1.96 * std.error)
  ) %>%
  select(term, OR, CI_low, CI_high, p.value)
```


```{r table4, echo=FALSE}
knitr::kable(table_or,  caption = "Association of risk factors with diabetes using Logistic Regression")

```


Interpretation 

The logistic regression in Table \@ref(tab:table4) identified several significant predictors of diagnosed diabetes. Age (OR=1.01, p<0.001) and BMI (OR=1.01, p=0.027) were positively associated with diabetes, indicating increased risk with higher values. Family history showed a strong effect (OR=1.50, p<0.001), highlighting genetic predisposition. HbA1c was the strongest predictor (OR=30.7, p<0.001), followed by fasting glucose (OR=1.02, p<0.001), reflecting their clinical importance. Higher physical activity (OR<1, p<0.001) and better diet score (borderline significance) were protective. HDL cholesterol was also inversely associated (p=0.029). Other variables, including hypertension, cardiovascular history, and blood pressure, were not statistically significant after adjustment.




## Predicted Probability of Diabetes (y) by HbA1c (x)

```{r pred_prob, echo=FALSE}

# Predicted probabilities dataset

newdata <- data.frame(
  hba1c = seq(min(data$hba1c, na.rm = TRUE),
              max(data$hba1c, na.rm = TRUE),
              length.out = 100),
  age = mean(data$age, na.rm = TRUE),
  bmi = mean(data$bmi, na.rm = TRUE),
  waist_to_hip_ratio =  mean(data$waist_to_hip_ratio, na.rm = TRUE),
  physical_activity_minutes_per_week = mean(data$physical_activity_minutes_per_week, na.rm = TRUE),
  diet_score = mean(data$diet_score, na.rm = TRUE),
  screen_time_hours_per_day = mean(data$screen_time_hours_per_day, na.rm = TRUE),
  glucose_fasting = mean(data$glucose_fasting, na.rm = TRUE),
  glucose_postprandial = mean(data$glucose_postprandial, na.rm = TRUE),
  systolic_bp = mean(data$systolic_bp, na.rm = TRUE),   
  diastolic_bp  = mean(data$diastolic_bp , na.rm = TRUE), 
  heart_rate = mean(data$heart_rate, na.rm = TRUE),   
  cholesterol_total   = mean(data$cholesterol_total  , na.rm = TRUE), 
  hdl_cholesterol   = mean(data$hdl_cholesterol, na.rm = TRUE),   
  ldl_cholesterol  = mean(data$ldl_cholesterol  , na.rm = TRUE), 
  triglycerides  = mean(data$triglycerides , na.rm = TRUE),   
  family_history_diabetes = factor("Yes", levels = c("No","Yes")),
  hypertension_history = factor("No", levels = c("No","Yes")),
  cardiovascular_history = factor("Yes", levels = c("No","Yes")),
  hypertension_history = factor("No", levels = c("No","Yes"))
)


newdata$predicted_prob <- predict(model, newdata, type = "response")
```

```{r chart4, echo=FALSE, fig.cap="Predicted Probability of Diabetes by HbA1c"}

#Plot


ggplot(newdata, aes(x = hba1c, y = predicted_prob)) +
  geom_line() +
  labs(
    title = "Predicted Probability of Diabetes by HbA1c",
    x = "HbA1c",
    y = "Predicted Probability"
  ) +
  theme_minimal()

```

Interpretation:

Chart \@ref(fig:chart4) shows that at low HbA1c values (around 4–5.5%), the predicted probability of diabetes is near zero, indicating minimal risk.
As HbA1c rises into the intermediate range (approximately 6–8%), the predicted probability increases sharply, indicating that HbA1c ≈ 6.5% as a critical threshold where individuals move from low to high risk. 
Beyond HbA1c values of 8, the predicted probability approaches 1, suggesting that nearly all individuals in this range are classified as diabetic.


# Conclusion


Overall, the findings indicate that age is the primary demographic factor associated with diabetes, while other socio-demographic characteristics show minimal differences between groups, consistent with evidence linking ageing to increased metabolic risk (World Health Organization, 2023). In contrast, lifestyle behaviours demonstrate a clearer relationship with diabetes risk, with low physical activity, poor diet, inadequate sleep, and higher screen time linked to elevated risk scores. However, screen time alone appears to be a weak independent predictor (Ekelund et al., 2019). These results suggest that while demographic factors provide limited differentiation, modifiable lifestyle behaviours play a more meaningful role in shaping diabetes risk. 

The logistic regression analysis reinforced the multifactorial nature of diabetes, integrating biomarkers, demographics, and lifestyle. HbA1c emerged as the most powerful predictor, with a sharp inflection around 6.5% marking the transition from low to high probability of diabetes. This threshold is consistent with established diagnostic guidelines and highlights HbA1c’s central role in risk stratification (American Diabetes Association, 2024). Fasting glucose also demonstrated significant predictive value, reinforcing its clinical utility in early detection (International Diabetes Federation, 2025). 

Age, BMI, and family history were positively associated with diabetes, reflecting some genetic influences. These findings align with prior epidemiological studies that emphasize the interplay of age-related metabolic decline, adiposity, and hereditary predisposition (Ruze R et al., 2023). Conversely, protective factors such as higher physical activity, better diet quality, and elevated HDL cholesterol were linked to reduced risk, underscoring the importance of lifestyle interventions in prevention (Ong et al., 2023; Park et al., 2020). 


# References

•	American Diabetes Association. (2024). Standards of medical care in diabetes—2024. Diabetes Care, 47(Suppl. 1), S1–S154. https://doi.org/10.2337/dc24-Sint


•	Dam, N., Verma, A., & Jasrotia, A. S. (2025). IDF atlas 2025: Global Diabetes Prevalence and Gestational Diabetes Mellitus . Diabetes Asia Journal, 2(1), 21–24. https://doi.org/10.62996/daj.54042025

• Ekelund, U., Tarp, J., Steene-Johannessen, J., Hansen, B. H., Jefferis, B., Fagerland, M. W., Whincup, P., Diaz, K. M., Hooker, S. P., Chernofsky, A., Larson, M. G., Spartano, N., Vasan, R. S., Dohrn, I.-M., Hagströmer, M., Edwardson, C., Yates, T., Shiroma, E., Anderssen, S. A., & Lee, I.-M. (2019). Dose-response associations between accelerometry measured physical activity and sedentary time and all-cause mortality: Systematic review and harmonised meta-analysis. BMJ, 366, l4570. https://doi.org/10.1136/bmj.l4570




•	International Diabetes Federation. (2025). IDF Diabetes Atlas (11th ed.). Brussels, Belgium: IDF. https://diabetesatlas.org


•	Ong, K., Stafford, L., McLaughlin, S., et al. (2023). Global, regional, and national burden of diabetes from 1990 to 2021, with projections of prevalence to 2050: A systematic analysis for the Global Burden of Disease Study 2021. The Lancet, 402(10385), 203–234. https://doi.org/10.1016/S0140-6736(23)01352-1 


•	Park, J. H., Moon, J. H., Kim, H. J., Kong, M. H., & Oh, Y. H. (2020). Sedentary lifestyle: Overview of updated evidence of potential health risks. Korean Journal of Family Medicine, 41(6), 365–373. https://doi.org/10.4082/kjfm.20.0165

•	Ruze R, Liu T, Zou X, Song J, Chen Y, Xu R, Yin X, Xu Q (2023). Obesity and type 2 diabetes mellitus: connections in epidemiology, pathogenesis, and treatments. Front Endocrinol (Lausanne), 14:1161521. doi: 10.3389/fendo.2023.1161521. PMID: 37152942; PMCID: PMC10161731.

•	Savvopoulos, S., Hatzikirou, H., & Jelinek, H. F. (2024). Comparative analysis of biomarkers in type 2 diabetes patients with and without comorbidities: Insights into the role of hypertension and cardiovascular disease. Biomarker Insights, 19, 11772719231222111. https://doi.org/10.1177/11772719231222111 

• World Health Organization. (2023). Diabetes. https://www.who.int/news-room/fact-sheets/detail/diabetes

